\ourlongtitle

### 2021-08-23 Bernoulli/glo-2020-pm-PaRIS.tex
Corrected typos

Thus, a pointwise estimate of $\ell(x)$ can be obtained by generating $\zeta$ from $\mathbf{R}(x, \rmd z)$ and computing the statistic $\ell \langle \zeta \rangle(x)$, the \emph{pseudo marginal}. In Monte Carlo methods, the measure to replace, when necessary, the true marginal $\ell$ by its pseudo marginal is referred to as \emph{pseudo marginalisation}. Interestingly, even though pseudo marginalisation is based on the plug-in principle, it preserves typically the consistency of an algorithm. In order to see the this, let $\bar{\mathsf{X}} \eqdef \mathsf{X} \times \mathsf{Z}$ and $\bar{\mathcal{X}} \eqdef \mathcal{X} \tensprod \mathcal{Z}$; then one may define an extended target distribution $\bar{\pi}(\rmd \bar{x}) \eqdef \bar{\lambda}(\rmd \bar{x}) / \bar{\lambda} \1_{\bar{\mathsf{X}}} = \bar{\lambda}(\rmd \bar{x}) / \lambda \1_{\mathsf{X}}$ on $(\bar{\mathsf{X}}, \bar{\mathcal{X}})$, where 

@@

which is tractable. Note that $\alpha$ is indeed a pseudo-marginal version of the exact acceptance probability $1 \wedge \ell (\xi^\ast) \rho(\xi^\ast, \xi_m) / \ell (\xi_m) \rho(\xi_m, \xi^\ast)$ obtained if $\ell$ would be known. Note that the auxiliary variable $\zeta$ enters $\alpha$ only through the estimates $\ell \langle \zeta^\ast \rangle (\xi^\ast)$ and $\ell \langle \zeta_m \rangle (\xi_m)$, and since the latter has already been computed at the previous iteration there is no need of recomputing this quantity. In the \emph{Monte-Carlo-within-Metropolis algorithm} (see again \cite{andrieu:robert:2009}) $\ell \langle z \rangle(x)$ is a pointwise importance sampling estimate of $\ell(x)$ based on a Monte Carlo sample $\zeta$ generated from $\mathbf{R}(x, \cdot)$. Alternatively, the extended distribution $\bar{\pi}$ can be sampled using rejection sampling or importance sampling, leading to consistent pseudo-marginal formulations of these algorithms as well.

In the present paper we will generalise the pseudo-marginal approach towards biased estimation by allowing the function 
\begin{equation} \label{eq:biased:pseudo:marginalisation}
    \ell^\precpar(x) \eqdef \int \ell \langle z \rangle(x) \, \mathbf{R}(x, \rmd z)
\end{equation}
on $\Xset$ to be different from $\ell$. Here $\varepsilon \geq 0$ is some accuracy parameter describing the distance between $\ell^\precpar$ and $\ell$. Such biased estimates appear naturally when the law $\pi$ is, \eg, governed by a diffusion process and the density $\ell$ is approximated on the basis of different discretisation schemes; see the next section. In that case, $\precpar$ plays the role of the discretisation step size. In \eqref{eq:biased:pseudo:marginalisation}, also the estimator $\ell \langle z \rangle(x)$ and the kernel $\mathbf{R}(x, \rmd z)$ may depend on $\precpar$, even though this is suppressed in the notation. By introducing the possibly unnormalised measure $\lambda^\varepsilon(\rmd x) \eqdef \ell^\precpar(\rmd x) \, \mu(\rmd x)$ on $\Xfd$ we may define the \emph{skew} target probability measure
$$
    \pi^\varepsilon(\rmd x) \eqdef \frac{\lambda^\precpar(\rmd x)}{\lambda^\precpar \1_\Xset}
$$
on $\Xfd$. In the biased case, generating a sample $(\xi^i, \zeta^i, \omega^i)_{i = 1}^N$ targeting \eqref{eq:extended:target} will, by \eqref{eq:biased:pseudo:marginalisation}, provide a sample $(\xi^i, \omega^i)_{i = 1}^N$ targeting $\pi^\varepsilon$ as a by-product. Thus, of outmost importance is to obtain control over the bias between $\pi$ and $\pi^\precpar$, which is possible under the assumption that there exists a constant $c \geq 0$ such that for all $h \in \bmf{\Xfd}$ and $\varepsilon$,   
\begin{equation} \label{eq:lipschitz:simple:case}
    \left| \lambda^\varepsilon h - \lambda h \right| \leq c \varepsilon \| h \|_\infty. 
\end{equation}
For instance, in the diffusion process case mentioned above, a condition of type \eqref{eq:lipschitz:simple:case} holds, as we will see in Section~\ref{sec:theoretical:results}, typically true in the case where the density is approximated using the \emph{Durham-Gallant estimator} \cite{durham:gallant:2002}. Using that for all $h \in \bmf{\Xfd}$,  
$$
    \pi^\precpar h - \pi h = \pi^\precpar h \left( 1 - \frac{\lambda^\varepsilon \1_\Xset}{\lambda \1_\Xset} \right) + \frac{\lambda^\precpar h - \lambda h}{\lambda \1_\Xset}, 
$$
we straightforwardly obtain the bound  
$$
    \left| \pi^\precpar h - \pi h \right| \leq \precpar \frac{2 c}{\lambda \1_\Xset} \| h \|_\infty
$$
on the systematic error induced by the skew model. Note that the unbiased case \eqref{eq:pseudo:marginalisation} corresponds to letting $\precpar = 0$ in assumption~\eqref{eq:lipschitz:simple:case}. In the next section we will present a solution to the main problem addressed in Section~\ref{sec:model} exploring a pseudo-marginalised version of the PaRIS discussed in Section~\ref{sec:PaRIS}.

@@

In the case where each $\ud{n}$ is intractable, so is the mixture distribution $\partmixt_n$ defined in \eqref{eq:def:partmixt}. Under \hypref{assum:biased:estimate}, we may, as in Section~\ref{sec:pseudo:marginalisation}, aim at consistent pseudo-marginalisation of the forward-sampling operation by applying self-normalised importance sampling to the extended mixture  
$$
\bar{\pi}_n(i, \rmd x, \rmd z)
\propto \ewght{n}{i} \ukest{n}{z}(\epart{n}{i}, x) \, \mu_{n + 1}(\rmd x) \, \ukdist{n}(\epart{n}{i}, x, \rmd z) 
$$
on $\bar{\Xfd}_{n + 1} \eqdef \powerset{\intvect{1}{\N}} \tensprod \Xfd_{n + 1} \tensprod \Zfd_{n + 1}$ using the instrumental distribution  
$$
\bar{\rho}_n(i, \rmd x, \rmd z) \propto \ewght{n}{i} \adjfuncforward{n}(\epart{n}{i}) \, \prop{n}(\epart{n}{i}, \rmd x) \, \ukdist{n}(\epart{n}{i}, x, \rmd z)
$$
on the same space. Note that the marginal of $\bar{\pi}_n$ with respect to $(i, x)$ is the distribution proportional to $\ewght{n}{i} \ukmod{n}(\epart{n}{i}, \rmd x)$, whose distance to the target $\pi_n$ of interest is controlled by the precision parameter $\precpar$. Here the adjustment multiplier $\adjfuncforward{n}$ and the proposal kernel $\prop{n}$ of the instrumental distribution are as in Section~\ref{sec:SMC}. Each draw from $\bar{\rho}_n$ is assigned an importance weight given by the (tractable) Radon--Nikodym derivative of $\bar{\pi}_n$ to $\bar{\rho}_n$. It is easily seen that this sampling operation, which is detailed in Algorithm~\ref{alg:pm:SMC}, corresponds to replacing the intractable transition density $\ud{n}$ on Line~4 in Algorithm~\ref{alg:ideal:SMC} by an estimate provided by \hypref{assum:biased:estimate}; we will hence refer to Algorithm~\ref{alg:pm:SMC} as \emph{pseudo-marginal forward sampling} and express it concisely as 
$$
    (\epart{n + 1}{i}, \ewght{n + 1}{i})_{i = 1}^\N \sim \mathsf{pmFS}((\epart{n}{i}, \ewght{n}{i})_{i = 1}^\N). 
$$

>>>>

Thus, a pointwise estimate of $\ell(x)$ can be obtained by generating $\zeta$ from $\mathbf{R}(x, \rmd z)$ and computing the statistic $\ell \langle \zeta \rangle(x)$, the \emph{pseudo marginal}. In Monte Carlo methods, the operation of replacing, when necessary, the true marginal $\ell$ by its pseudo marginal is referred to as \emph{pseudo marginalisation}. Interestingly, even though pseudo marginalisation is based on the plug-in principle, it preserves typically the consistency of an algorithm. In order to see the this, let $\bar{\mathsf{X}} \eqdef \mathsf{X} \times \mathsf{Z}$ and $\bar{\mathcal{X}} \eqdef \mathcal{X} \tensprod \mathcal{Z}$; then one may define an extended target distribution $\bar{\pi}(\rmd \bar{x}) \eqdef \bar{\lambda}(\rmd \bar{x}) / \bar{\lambda} \1_{\bar{\mathsf{X}}} = \bar{\lambda}(\rmd \bar{x}) / \lambda \1_{\mathsf{X}}$ on $(\bar{\mathsf{X}}, \bar{\mathcal{X}})$, where 

@@

which is tractable. Note that $\alpha$ is indeed a pseudo-marginal version of the exact acceptance probability $1 \wedge \ell (\xi^\ast) \rho(\xi^\ast, \xi_m) / \ell (\xi_m) \rho(\xi_m, \xi^\ast)$ obtained if $\ell$ was known. Note that the auxiliary variable $\zeta$ enters $\alpha$ only through the estimates $\ell \langle \zeta^\ast \rangle (\xi^\ast)$ and $\ell \langle \zeta_m \rangle (\xi_m)$, and since the latter has already been computed at the previous iteration there is no need of recomputing this quantity. In the \emph{Monte-Carlo-within-Metropolis algorithm} (see again \cite{andrieu:robert:2009}) $\ell \langle z \rangle(x)$ is a pointwise importance sampling estimate of $\ell(x)$ based on a Monte Carlo sample $\zeta$ generated from $\mathbf{R}(x, \cdot)$. Alternatively, the extended distribution $\bar{\pi}$ can be sampled using rejection sampling or importance sampling, leading to consistent pseudo-marginal formulations of these algorithms as well.

In the present paper we will generalise the pseudo-marginal approach towards biased estimation by allowing the function 
\begin{equation} \label{eq:biased:pseudo:marginalisation}
    \ell^\precpar(x) \eqdef \int \ell \langle z \rangle(x) \, \mathbf{R}(x, \rmd z)
\end{equation}
on $\Xset$ to be different from $\ell$. Here $\varepsilon \geq 0$ is some accuracy parameter describing the distance between $\ell^\precpar$ and $\ell$. Such biased estimates appear naturally, e.g. when the law $\pi$ is governed by a diffusion process and the density $\ell$ is approximated on the basis of different discretisation schemes; see the next section. In that case, $\precpar$ plays the role of the discretisation step size. In \eqref{eq:biased:pseudo:marginalisation}, also the estimator $\ell \langle z \rangle(x)$ and the kernel $\mathbf{R}(x, \rmd z)$ may depend on $\precpar$, even though this is suppressed in the notation. By introducing the possibly unnormalised measure $\lambda^\varepsilon(\rmd x) \eqdef \ell^\precpar(\rmd x) \, \mu(\rmd x)$ on $\Xfd$ we may define the \emph{skew} target probability measure
$$
    \pi^\varepsilon(\rmd x) \eqdef \frac{\lambda^\precpar(\rmd x)}{\lambda^\precpar \1_\Xset}
$$
on $\Xfd$. In the biased case, generating a sample $(\xi^i, \zeta^i, \omega^i)_{i = 1}^N$ targeting \eqref{eq:extended:target} will, by \eqref{eq:biased:pseudo:marginalisation}, provide a sample $(\xi^i, \omega^i)_{i = 1}^N$ targeting $\pi^\varepsilon$ as a by-product. Thus, it is of utmost importance to obtain control over the bias between $\pi$ and $\pi^\precpar$, which is possible under the assumption that there exists a constant $c \geq 0$ such that for all $h \in \bmf{\Xfd}$ and $\varepsilon$,   
\begin{equation} \label{eq:lipschitz:simple:case}
    \left| \lambda^\varepsilon h - \lambda h \right| \leq c \varepsilon \| h \|_\infty. 
\end{equation}
For instance, in the diffusion process case mentioned above, a condition of type \eqref{eq:lipschitz:simple:case} holds, as we will see in Section~\ref{sec:theoretical:results}, typically true in the case where the density is approximated using the \emph{Durham-Gallant estimator} \cite{durham:gallant:2002}. Using that for all $h \in \bmf{\Xfd}$,  
$$
    \pi^\precpar h - \pi h = \pi^\precpar h \left( 1 - \frac{\lambda^\varepsilon \1_\Xset}{\lambda \1_\Xset} \right) + \frac{\lambda^\precpar h - \lambda h}{\lambda \1_\Xset}, 
$$
we straightforwardly obtain the bound  
$$
    \left| \pi^\precpar h - \pi h \right| \leq \precpar \frac{2 c}{\lambda \1_\Xset} \| h \|_\infty
$$
on the systematic error induced by the skew model. Note that the unbiased case \eqref{eq:pseudo:marginalisation} corresponds to letting $\precpar = 0$ in assumption~\eqref{eq:lipschitz:simple:case}. In the next section we will present a solution to the main problem addressed in Section~\ref{sec:model} exploring a pseudo-marginalised version of the PaRIS discussed in Section~\ref{sec:PaRIS}.

@@

In the case where each $\ud{n}$ is intractable, so is the mixture distribution $\partmixt_n$ defined in \eqref{eq:def:partmixt}. Under \hypref{assum:biased:estimate}, we may, as in Section~\ref{sec:pseudo:marginalisation}, aim at consistent pseudo-marginalisation of the forward-sampling operation by applying self-normalised importance sampling to the extended mixture  
$$
\bar{\pi}_n(i, \rmd x, \rmd z)
\propto \ewght{n}{i} \ukest{n}{z}(\epart{n}{i}, x) \, \mu_{n + 1}(\rmd x) \, \ukdist{n}(\epart{n}{i}, x, \rmd z) 
$$
on $\bar{\Xfd}_{n + 1} \eqdef \powerset{\intvect{1}{\N}} \tensprod \Xfd_{n + 1} \tensprod \Zfd_{n + 1}$ using the instrumental distribution  
$$
\bar{\rho}_n(i, \rmd x, \rmd z) \propto \ewght{n}{i} \adjfuncforward{n}(\epart{n}{i}) \, \prop{n}(\epart{n}{i}, \rmd x) \, \ukdist{n}(\epart{n}{i}, x, \rmd z)
$$
on the same space. Note that the marginal of $\bar{\pi}_n$ with respect to $(i, x)$ is the distribution proportional to $\ewght{n}{i} \ukmod{n}(\epart{n}{i}, \rmd x)$, whose distance to the target $\pi_n$ of interest is controlled by the precision parameter $\precpar$. Here the adjustment multiplier $\adjfuncforward{n}$ and the proposal kernel $\prop{n}$ of the instrumental distribution are as in Section~\ref{sec:SMC}. Each draw from $\bar{\rho}_n$ is assigned an importance weight given by the (tractable) Radon--Nikodym derivative of $\bar{\pi}_n$ with respect to $\bar{\rho}_n$. It is easily seen that this sampling operation, which is detailed in Algorithm~\ref{alg:pm:SMC}, corresponds to replacing the intractable transition density $\ud{n}$ on Line~4 in Algorithm~\ref{alg:ideal:SMC} by an estimate provided by \hypref{assum:biased:estimate}; we will hence refer to Algorithm~\ref{alg:pm:SMC} as \emph{pseudo-marginal forward sampling} and express it concisely as 
$$
    (\epart{n + 1}{i}, \ewght{n + 1}{i})_{i = 1}^\N \sim \mathsf{pmFS}((\epart{n}{i}, \ewght{n}{i})_{i = 1}^\N). 
$$
